---
title: "5.01 Assimilate Penn's Landing WQ Time Series Data"
author: "J. Hagy"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("wql")
#install.packages('dataRetrieval')

#https://rdrr.io/cran/wql/
# Exploring Water Quality Monitoring Data - Jassby, Cloern, Stachelek

library(tidyverse)
library(lubridate)
library(marelac)
library(wql)
library(wesanderson)
library(viridis)
library(here)

large_text <- theme(axis.text=element_text(size=14),
        axis.title=element_text(size=14))
no_grid <- theme(panel.grid=element_blank())

pal <- wes_palette("Zissou1",5,type="discrete")

# Get function to impute missing values
source(here("functions","imputeWQbyYear.R"))

```

## Download Penn's Landing WQ Data
This code chunk creates the data frame drPenn_ctdo which has variables for daily mean conductivity (µS/cm), water temperature (°C), and dissolved oxygen (mg/L) for 1961-12-27 through 20213-1-30.  An Rdata file was saved when the code was originally created but is not overwritten to preserve data integrity.

```{r read data, echo=TRUE}

con <- file(here("raw","PennsLandingDV (retrieved 2023-03-09).txt"))
tmp <- readLines(con,n=52) %>% .[51] %>% strsplit(.,split="\t") %>% .[[1]]

drPenn <- read.table(file=here("raw","PennsLandingDV (retrieved 2023-03-09).txt"),
                     col.names=tmp,skip=52,sep="\t")

# The monitoring data at Penn's Landing includes data from two nearby but not
# identical stations, which here are called the "original deployment" and the
# "ISM Barge".  USGS recorded these data as different variables, so they are
# precessed separately and then combined for the purposes of this study.

## Data set 1 includes data from original deployment
# Subset mean spCond,wt,DO(mg/L)
drPenn_ctdo.1 <- drPenn[,c(1,2,3,8,14,20)]
names(drPenn_ctdo.1) <- c("agency_cd","site_no","date","spCond.mean","wt.mean","do.mgL.mean")
drPenn_ctdo.1 <- drPenn_ctdo.1 %>% mutate(date=as.Date(date))
summary(drPenn_ctdo.1)

## Data set 2 includes data from ISM barge 
# Subset mean spCond,wt,DO(mg/L) from ISM Test Bed [ISM Barge]
drPenn_ctdo.2 <- drPenn %>% .[,c(1,2,3,28,44,30)]
names(drPenn_ctdo.2) <- c("agency_cd","site_no","date","spCond.mean","wt.mean","do.mgL.mean")
drPenn_ctdo.2 <- drPenn_ctdo.2 %>% mutate(date=as.Date(date))
summary(drPenn_ctdo.2)

# Graph the two data sets to show that data set 1 goes to near the end of 2020
# while data set 2 begins near the middle of 2020.
rbind(
  drPenn_ctdo.1 %>% 
  pivot_longer(cols=all_of(c("spCond.mean","wt.mean","do.mgL.mean")),
              names_to="param",values_to="value") %>%
  filter(year(date)>2009) %>% 
  mutate(loc="Original Deployment"),
  drPenn_ctdo.2 %>% 
  pivot_longer(cols=all_of(c("spCond.mean","wt.mean","do.mgL.mean")),
              names_to="param",values_to="value") %>%
  filter(year(date)>2009) %>% 
  mutate(loc="ISM Barge")
) %>% 
ggplot(aes(x=date,y=value,color=loc))+
    facet_wrap(~param,scales="free_y",ncol=1)+
    geom_point(shape=1,alpha=0.5)+
    no_grid

# To combine the data, use data from data set 1 through the end of 2020
# and use data from data set 2 for 2021 and later.
drPenn_ctdo <- rbind(
  drPenn_ctdo.1 %>% filter(year(date)<2021),
  drPenn_ctdo.2 %>% filter(year(date)>=2021))

# Check that the dataset is now a continuous time series
drPenn_ctdo %>% 
  pivot_longer(cols=all_of(c("spCond.mean","wt.mean","do.mgL.mean")),
              names_to="param",values_to="value") %>%
  filter(year(date)>2009) %>% 
  ggplot(aes(x=date,y=value))+
    facet_wrap(~param,scales="free_y",ncol=1)+
    geom_point()+
  no_grid

# Save the data as an Rdata file
#save(drPenn,drPenn_ctdo,file=here("data","5.01 Delaware River at Penn Landing.Rdata"))

```

## Review the Penn Landing Data 
Ensure data were read correctly and are sufficient for the 
intended analysis 

```{r review data, echo=TRUE, warning=FALSE}
# Load the Penn's Landing Time Series
load(here("data","5.01 Delaware River at Penn Landing.Rdata"))

# Examine statistics to make sure data was read in correctly
# Ranges shown here are appropriate
summary(drPenn_ctdo)

# Plot the Penn Landing data since 2002
drPenn_ctdo %>% 
  pivot_longer(cols=all_of(c("spCond.mean","wt.mean","do.mgL.mean")),names_to="param",values_to="value") %>% 
  filter(year(date)>2001) %>% 
  ggplot(aes(x=date,y=value))+
    facet_wrap(~param,scales="free_y",ncol=1)+
    geom_point()+
    scale_x_date(date_breaks="2 years",date_labels="%Y")+
    theme(panel.grid=element_blank())

# Calculate salinity using wql::ec2pss.  This functions required
# specific conductivity as mS/cm, but USGS provides spCond as 
# µS/cm.  Therefore, divide by 1000 before calculating salinity 
tmp <- drPenn_ctdo %>% 
  mutate(year=year(date)) %>% 
  select(date,year,wt.mean,do.mgL.mean,spCond.mean) %>% 
  filter(complete.cases(.)) %>% 

  # Calculate percent oxygen saturation.
  # If salinity is less than 1, use APHA method that does not use salinity. APHA
  # method is the standard for fresh water application.
  # If salinity is >=1, then use "Weiss" method, which considers salinity, and is 
  # the standard formula in marine sciences (Soetert and Petzoldt 2020).

  # Soetaert K, Petzoldt T (2020). _marelac: Tools for Aquatic Sciences_. R package
  # version 2.1.10, <https://CRAN.R-project.org/package=marelac>.
  mutate(salinity.mean=ec2pss(ec=spCond.mean/1000,t=wt.mean),
      posat.mean=ifelse(salinity.mean<1,
      do.mgL.mean*100/gas_O2sat(t=wt.mean,method="APHA"),
      do.mgL.mean*100/gas_O2sat(S=salinity.mean,t=wt.mean,method="Weiss")
      ))
# Show that salinity and percent DO saturation are calculated correctly
tmp %>% 
  pivot_longer(cols=all_of(c("salinity.mean","wt.mean","posat.mean")),
               names_to="param",values_to="value") %>% 
  filter(year(date)>2008) %>% 
  ggplot(aes(x=date,y=value))+
    facet_wrap(~param,scales="free_y",ncol=1)+
    geom_point()+
  theme(panel.grid=element_blank())

# Count observations by year/month.
# A lot of the data from December-February are absent.  
# This is not a concern since modeling is July 1-November 1.
# From 2019 onward, the data is year-round
drPenn_ctdo %>% 
  filter(complete.cases(.)) %>% 
  mutate(month=month(date),year=year(date)) %>% 
  group_by(year,month) %>% 
  summarize(n=n()) %>% 
  filter(year>2008) %>% 
  ggplot(aes(x=month,y=n))+
  facet_wrap(~year,ncol=2)+
  geom_col()+
  scale_x_continuous(breaks=seq(1,12))
```

## Plot POSAT for 20 years at Penn's Landing
Create a facet plot showing each year of DO data from 2002 to 2022. This figure appears in the Technical Report

```{r Plot for Tech Report, echo=TRUE}

## Plot DO with water temperature
# Plot 20 years of DO at Penn Landing.
# recast the dates as a date in 2002 so that same scale can be 
# used in all the facets.  Dotted lines show Jul 1-Oct 31.
plt <- tmp %>% 
  filter(between(year(date),2002,2022)) %>%
  mutate(dy=as.Date(paste0("2002-",format(date,format="%m-%d")))) %>% 
  ggplot(aes(x=dy,y=posat.mean,color=wt.mean))+
  geom_point()+
  facet_wrap(~year(date),ncol=4)+
  scale_color_continuous(type="viridis",breaks=seq(5,25,by=5))+
  scale_x_date(
    limits=c(as.Date("2002-06-01"),as.Date("2002-11-30")),
    breaks=as.Date(c("2002-06-01","2002-07-01","2002-08-01",
        "2002-09-01","2002-10-01","2002-11-01")),
    labels=c("","7/1","","9/1","","11/1"))+
  labs(x="",y="Percent Oxygen Saturation")+
  ggtitle("Penn's Landing, PA Dissolved Oxygen, 2002-2022")+
  geom_vline(xintercept=c(as.Date("2002-07-01"),as.Date("2002-10-31")),linetype=2)+
  no_grid+
  theme(legend.position="inside",legend.position.inside=c(0.6,0.05),legend.direction="horizontal")+
  guides(color=guide_colorbar(title="Water Temperature"))
plt

# This creates figure A1-2 in the TSD.
 pdf(file=here("figures","5.01 Penn Landing POSAT 2002-2022.pdf"),
     width=6,height=8,useDingbats = FALSE,paper="letter")
   plt
 dev.off()

```

## Generate Summer Complete Data
This code chunk uses linear interpolation to impute values for any missing data between July 1 and November 30.

```{r Impute Missing Data, echo=TRUE}

# Make a summer time series for each year from 2002 to 2022 excluding 2010 which
# we believe has too much missing data to define the time series adequately to
# run #simulations.
Year <- c(seq(2002,2009,by=1),seq(2011,2022,by=1))
# Define data to be imputed
ts_input <- drPenn_ctdo %>% mutate(year=year(date))
# Impute the data
drpl_summer <- do.call(rbind,lapply(Year,imputeByYear))
drpl_summer <- filter(drpl_summer,!is.na(year))

# Don't overwrite file as saved.
save(drpl_summer,file=here("data","5.01 DR_Penn_Landing_Summer_Complete 2002-2022.Rdata"))

```

