---
title: "6.03 Fit QGAMs to HSI vs. POSAT Percentiles"
author: "J. Hagy"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(marelac)
library(wesanderson)
library(viridis)
library(qgam)

large_text <- theme(axis.text=element_text(size=14),
        axis.title=element_text(size=14),
        strip.text=element_text(size=14))
no_grid <- theme(panel.grid=element_blank())

pal <- wes_palette("Zissou1",5,type="discrete")

dataPath <- "data/"
figuresPath <- "figures/"
outputsPath <- "outputs/"
```

# Load Percentiles and HSI data

```{r Load Data}

# Load Percentiles
load(paste0(dataPath,"6.02 DO and POSAT Percentiles.Rdata"))

# Load HSI Data
load(paste0(dataPath,"5.03 Chester Cohort Simulation 2002-2022.Rdata"))
load(paste0(dataPath,"5.04 Penns Landing Cohort Simulation 2002-2022.Rdata"))

load(paste0(dataPath,"6.01 Simulated HSI using Restored DO.Rdata"))
# confirmed this has exactly the same HSI values as original program 5.72 

hsi.observed <- rbind(
  sim.Chester.Zmin %>% mutate(Site="Chester",type="Observed"),
  sim.PennsLanding.Zmin %>% mutate(Site="Penn's Landing",type="Observed")) %>% 
  mutate(year=as.numeric(year)) %>% 
  filter(between(doy,182,304)) %>% 
  group_by(Site,type,year) %>% 
  summarize(mean.growth=mean(growth),mean.Zmin=mean(mortality)) %>% 
  mutate(mean.phi=mean.growth-mean.Zmin) %>% 
  ungroup()

hsi.restored <- restoredHSI %>% mutate(type="Restored")

hsi <- rbind(hsi.observed,hsi.restored)

```

## Calculate descriptive statistics for HSI.
These values appear in the TSD on page 32.

```{r}
# Descriptive statistics for HSI
hsi %>% filter(type=="Observed") %>% select(mean.phi) %>% summary()
hsi %>% filter(type=="Restored") %>% select(mean.phi) %>% summary()

# Descriptive statistics for observed DO.
# HSI varied from -0.082 at Penn's Landing in 2005 to 0.0056 at
# Chester in 2018. Biomass change varied from nearly 100% loss (<0.01% remaining)
# to a nearly 2-fold increase.
tmp <- hsi.observed %>% 
  mutate(biomass.change = exp(mean.phi*123))
tmp <- tmp[order(tmp$mean.phi),]
tmp[c(1,41),]

hsi.observed %>% 
  group_by(Site) %>% 
  summarize(mean.phi = mean(mean.phi))

fit <- lm(mean.phi ~ year, data=hsi.observed) 
summary(fit)

hsi %>% group_by(type) %>% 
  summarize(hsi.mean=mean(mean.phi))

```

# Combine Percentiles and HSI

```{r Combine Percentiles and HSI}

#Join HSI data with percentiles
df.obs <- left_join(hsi,percentiles,by=c("Site","type","year")) %>% 
  filter(year != 2010) %>% 
  filter(percentile %in% c(0.1,0.5))

```

## Fit QGAMs for POSAT and Calculate Predicted HSI

```{r Fit qGAMs}

# Create a function to predict QGAMs
fitQGAM.POSAT <- function(pct=NULL,typ=NULL,tau=0.5) {
  if(typ=="Combined") {
  df.subset <- df.obs %>% 
    filter(percentile==pct)
  } else {
  df.subset <- df.obs %>% 
  filter(percentile==pct & type==typ)
  }
  fit <- qgam(mean.phi ~ s(posat,k=3),qu=tau,data=df.subset)
  return(fit)
}

# Fit QGAMs to Observed, Restored, and Combined Data
fit.10.obs <- fitQGAM.POSAT(0.10,"Observed")
fit.50.obs <- fitQGAM.POSAT(0.50,"Observed")
fit.10.res <- fitQGAM.POSAT(0.10,"Restored")
fit.50.res <- fitQGAM.POSAT(0.50,"Restored")
fit.10.com <- fitQGAM.POSAT(0.10,"Combined")
fit.50.com <- fitQGAM.POSAT(0.50,"Combined")

# Model summary for 10th Percentile Observed
summary(fit.10.obs)
check(fit.10.obs) # check for convergence

# Model summary for Median Observed
summary(fit.50.obs)
check(fit.50.obs) # check for convergence

# Model summary for 10th Percentile Restored
summary(fit.10.res)
check(fit.10.res) # check for convergence

# Model summary for Median Restored
summary(fit.50.res)
check(fit.50.res) # check for convergence

# Model summary for 10th Percentile Combined Data
summary(fit.10.com)
check(fit.10.com) # check for convergence

# Model summary for Median Combined Data
summary(fit.50.com)
check(fit.50.com) # check for convergence

## Create a function to use QGAM models to predict PP at a range of percentiles
qgamPredict <- function(inputModel,pct=NULL,typ=NULL) {
  if(typ=="Combined") {
    df.subset <- df.obs %>% filter(percentile==pct)
  } else {
    df.subset <- df.obs %>% filter(type==typ & percentile==pct)
  }
  # create a template that spans the range of observations for the type and percentile
  # of interest
  df.pred <- data.frame(posat=seq(round(min(df.subset$posat),2),
                               round(max(df.subset$posat),2),
                               by=0.01))
  tmp <- predict(inputModel,df.pred,se.fit=TRUE)
  df.pred$pp.pred <- tmp$fit
  df.pred$se <- tmp$se.fit
  df.pred <- df.pred %>% mutate(percentile=pct,type=typ)
  return(df.pred)
}

df.pred <- rbind(
  qgamPredict(fit.10.obs,pct=0.1,typ="Observed"),
  qgamPredict(fit.50.obs,pct=0.5,typ="Observed"),
  qgamPredict(fit.10.res,pct=0.1,typ="Restored"),
  qgamPredict(fit.50.res,pct=0.5,typ="Restored"),
  qgamPredict(fit.10.com,pct=0.1,typ="Combined"),
  qgamPredict(fit.50.com,pct=0.5,typ="Combined")
)

# save predictions
q.predictions.POSAT <- df.pred
save(q.predictions.POSAT,file=paste0(dataPath,"6.03 Predicted PP and POSAT percentiles.Rdata"))

# Export data for CSV
write.csv(df.pred,file=paste0(outputsPath,"6.03 Predicted PP and POSAT percentiles.csv"))

```

# Calculate lowest POSAT with HSI>0
This code chunk calculates the POSAT thresholds for HSI>0 that appear in Table 3 of the TSD. 
```{r }

# Calculate and display the lowest DO percentile associated with HSI>0
crit <- df.pred %>% 
  filter(pp.pred>=0) %>% 
  group_by(type,percentile) %>% 
  summarize(crit=min(posat)) %>% 
  .[c(3,4,5,6,1,2),]
crit.ucl <- df.pred %>% 
  mutate(hsi.ll=pp.pred-se) %>% 
  filter(hsi.ll>0) %>% 
  group_by(type,percentile) %>% 
  summarize(crit.ucl=min(posat))
crit.lcl <- df.pred %>% 
  mutate(hsi.ul=pp.pred+se) %>% 
  filter(hsi.ul>0) %>% 
  group_by(type,percentile) %>% 
  summarize(crit.lcl=min(posat))
crit.all <- full_join(crit.ucl,crit.lcl,by=c("type","percentile")) %>% 
  full_join(crit,by=c("type","percentile")) %>% 
  .[c(5,6,3,4,1,2),c(1,2,5,4,3)]
crit.all
```

# Graph POSAT vs. HSI with Thresholds

```{r Graph POSAT vs HSI}

# Graph the POSAT vs. HSI values and associated QGAM model fit
# for Observed and Restored Seperately
pred.OandR <- df.pred %>% filter(type != "Combined")
plt <- ggplot()+
  geom_ribbon(data=pred.OandR,aes(x=posat,ymin=pp.pred-se,
    ymax=pp.pred+se),alpha=0.2)+
  geom_point(data=df.obs,aes(x=posat,y=mean.phi,color=Site))+
  geom_line(data=pred.OandR,aes(x=posat,y=pp.pred))+
  facet_grid(type~percentile,scales="free")+
  geom_hline(yintercept=0,linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.1 & type=="Observed"),
             aes(xintercept=crit$crit[1]),linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.5 & type=="Observed"),
             aes(xintercept=crit$crit[2]),linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.1 & type=="Restored"),
             aes(xintercept=crit$crit[3]),linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.5 & type=="Restored"),
             aes(xintercept=crit$crit[4]),linetype=2)+
  labs(x="Percent Oxygen Saturation",y="HSI")+
  no_grid+
  theme(legend.position="top")+
  guides(color=guide_legend(title="Site"))
plt

# This generates Figure 10 in the TSD. 
pdf(file=paste0(figuresPath,"6.03 HSI vs. POSAT quantiles.pdf"),
    width=6,height=6,paper="letter")
  plt
dev.off()

# Graph the POSAT vs. HSI values and associated QGAM model fit
# for Observed and Restored Combined
pred.OandR <- df.pred %>% filter(type == "Combined")
plt <- ggplot()+
  geom_ribbon(data=pred.OandR,aes(x=posat,ymin=pp.pred-se,
    ymax=pp.pred+se),alpha=0.2)+
  geom_point(data=df.obs,aes(x=posat,y=mean.phi,color=Site))+
  geom_line(data=pred.OandR,aes(x=posat,y=pp.pred))+
  facet_wrap(~percentile,scales="free_x",ncol=2)+
  geom_hline(yintercept=0,linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.1),
             aes(xintercept=crit$crit[5]),linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.5),
             aes(xintercept=crit$crit[6]),linetype=2)+
  labs(x="Percent Oxygen Saturation",y="HSI")+
  no_grid+
  theme(legend.position="top")+
  guides(color=guide_legend(title="Site"))
plt

# This generates Figure 11 in the TSD. 
pdf(file=paste0(figuresPath,"6.03 HSI vs. POSAT Combined quantiles.pdf"),
    width=6,height=4,paper="letter")
  plt
dev.off()

```

# Make "Zoomed-in" figure

```{r eval=FALSE, include=FALSE}

# This code does not contribute to the TSD.

# Graph the POSAT vs. HSI values and associated QGAM model fit
# for Observed and Restored Combined
pred.OandR <- df.pred %>% filter(type == "Combined")
plt <- ggplot()+
  geom_ribbon(data=pred.OandR,aes(x=posat,ymin=pp.pred-se,
    ymax=pp.pred+se),alpha=0.2)+
  geom_point(data=df.obs,aes(x=posat,y=mean.phi,color=Site))+
  geom_line(data=pred.OandR,aes(x=posat,y=pp.pred))+
  facet_wrap(~percentile,scales="free_x",ncol=2)+
  coord_cartesian(xlim=c(62,80),ylim=c(-0.01,0.01))+
  scale_x_continuous(breaks=seq(60,84,by=2))+
  scale_y_continuous(breaks=seq(-0.01,0.01,by=0.002))+
  geom_hline(yintercept=0,linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.1),
             aes(xintercept=crit$crit[5]),linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.5),
             aes(xintercept=crit$crit[6]),linetype=2)+
  labs(x="Percent Oxygen Saturation",y="HSI")+
  no_grid+
  theme(legend.position="top")+
  guides(color=guide_legend(title="Site"))
plt

pdf(file=paste0(figuresPath,"6.03 HSI vs. DO Combined quantiles (zoomed).pdf"),
    width=6,height=4,paper="letter")
  plt
dev.off()

# Graph the POSAT vs. HSI values and associated QGAM model fit
# for Observed and Restored Combined
pred.OandR <- df.pred %>% filter(type == "Observed" & percentile==0.5)
plt <- ggplot()+
  geom_ribbon(data=pred.OandR,aes(x=posat,ymin=pp.pred-se,
    ymax=pp.pred+se),alpha=0.2)+
  geom_point(data=filter(df.obs,percentile==0.5),aes(x=posat,y=mean.phi,color=Site))+
  geom_line(data=pred.OandR,aes(x=posat,y=pp.pred))+
  #facet_wrap(~percentile,scales="free_x",ncol=2)+
  coord_cartesian(xlim=c(66,85),ylim=c(-0.01,0.01))+
  scale_x_continuous(breaks=seq(66,84,by=2))+
  scale_y_continuous(breaks=seq(-0.01,0.01,by=0.002))+
  geom_hline(yintercept=0,linetype=2)+
  #geom_vline(data=subset(df.obs,percentile==0.1),
  #           aes(xintercept=crit$crit[5]),linetype=2)+
  geom_vline(data=subset(df.obs,percentile==0.5),
             aes(xintercept=crit$crit[2]),linetype=2)+
  labs(x="Percent Oxygen Saturation",y="Habitat Suitability Index")+
  no_grid+
  theme(legend.position="top")+
  guides(color=guide_legend(title="Site"))
plt

png(file=paste0(figuresPath,"6.03 HSI vs. DO Obs-Med quantiles (zoomed).png"),
    width=5,height=3,units="in",res=150)
  plt
dev.off()

