---
title: "7.01 Evaluate Seasonal Attainment"
author: "J. Hagy"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(marelac)
library(wql)
library(wesanderson)
library(viridis)
library(quantreg)


basic_theme <- theme_classic() +
  theme(axis.text=element_text(size=14),
        axis.title=element_text(size=14))
nogrid <- theme(panel.grid=element_blank())
largetext <- theme(axis.text=element_text(size=14),
                   axis.title = element_text(size=14))

pal <- wes_palette("Zissou1",5,type="discrete")

dataPath <- "data/"
outputsPath <- "outputs/"

```

## Evaluate Seasonal Attainment
These calculations are used in the economic analysis to determine whether additional wastewater treatment plant controls are needed to attain EPA’s proposed criteria in each of the three seasons. 

## Load the data from Chester and Penn's Landing and Calculate daily stats

```{r loadData}

load(paste0(dataPath,"5. Chester_1961-2023.Rdata"))
load(paste0(dataPath,"5.01 Delaware River at Penn Landing.Rdata"))

df <- rbind(
  drChester_ctdo %>% mutate(site="Chester"),
  drPenn_ctdo %>% mutate(site="Penn's Landing")) %>% 
  filter(date>as.Date("2011-11-01")) %>%
  filter(date<as.Date("2022-10-31")) %>% 
  select(date,site,wt.mean,do.mgL.mean,spCond.mean) %>% 
  filter(complete.cases(.)) %>% 
  mutate(salinity.mean=ec2pss(ec=spCond.mean/1000,t=wt.mean),
         posat.mean = do.mgL.mean*100/gas_O2sat(S=salinity.mean,t=wt.mean)) %>% 
  select(-spCond.mean)

df.long <- df %>% pivot_longer(cols=!all_of(c("site","date")),names_to = "param",values_to = "value")

daily_stats <- df.long %>% 
  mutate(doy=yday(date)) %>% 
  group_by(site,param,doy) %>% 
  summarize(mn=min(value),med=median(value),mx=max(value)) %>% 
  ungroup() %>% 
  filter(param != "salinity.mean") %>% 
  mutate(season=cut(doy,breaks=c(0,60,182,304,366),
             labels=c("Overwinter","Spawning","Growth","Overwinter"))) 

# Calculate the seasonal temperatures used in proposed rule for Alternative 1 (criteria expressed as concentration).  Values included in rule are for Chester only.
daily_stats %>% 
  group_by(site,param,season) %>% 
  summarize(p.10=quantile(med,probs=0.1),
            mean=mean(med),
            p.90=quantile(med,probs=0.9)) %>% 
  mutate(crit.50=0.66*gas_O2sat(S=0,t=mean),
         crit.90=0.66*gas_O2sat(S=0,t=p.90))
```

# Are the Spawning and Overwintering Criteria Currently Attained at Chester
This information is used in the economic analysis for the rule.

```{r MakeAssessment}

tmp <- df.long %>% 
  filter(param=="posat.mean") %>%
  mutate(doy=yday(date)) %>% 
  mutate(season=cut(doy,breaks=c(0,60,182,304,366),
             labels=c("Overwinter","Spawning","Growth","Overwinter"))) %>% 
  mutate(year=year(date)) %>% 
  mutate(assess_year=ifelse(yday(date)>305,year+1,year)) %>% 
  mutate(posat_lt_66=value<66,
         posat_lt_74=value<74)

tmp.1 <- tmp %>% 
  group_by(site,season,assess_year) %>% 
  summarize(n=n(),
            p.10=quantile(value,probs=0.1),
            p.50=quantile(value,probs=0.5))

tmp.2 <-  tmp %>% 
  filter(posat_lt_66) %>% 
  group_by(site,season,assess_year) %>% 
  summarize(days_lt_66=n())

tmp.3 <-  tmp %>% 
  filter(season=="Growth") %>% 
  filter(posat_lt_74) %>% 
  group_by(site,season,assess_year) %>% 
  summarize(days_lt_74=n()) 

assess <- tmp.1 %>%
  select(assess_year,season,n,p.10,p.50) %>% 
  left_join(.,tmp.2,by=c("site","season","assess_year")) %>% 
  left_join(.,tmp.3,by=c("site","season","assess_year"))

assess$days_lt_74[is.na(assess$days_lt_74) & assess$season=="Growth"] <- 0
assess$days_lt_66[is.na(assess$days_lt_66)] <- 0

assess <- assess %>% 
  mutate(
    attains.10=days_lt_66<=12,
    attains.50=days_lt_74<=61,
    attains=ifelse(season=="Growth",attains.10 & attains.50,attains.10)
  )

write.csv(assess,file=paste0(outputsPath,"7.01 Seasonal Criteria Assessment Analysis.csv"))

```

## Make seasonal graphs

```{r PlotSeasonalGraphs}

# add date, assuming year=2019 (but not showing it)
daily_stats <- daily_stats %>% mutate(date=as.Date(doy,origin="2018-12-31"))
ggplot()+
  geom_line(data=daily_stats,aes(x=date,y=med),color=pal[1])+
  geom_ribbon(data=daily_stats,aes(x=date,ymin=mn,ymax=mx),fill=pal[2],alpha=0.5)+
  facet_wrap(~param,ncol=1,scales="free_y")+
  scale_x_date(date_breaks = "1 month",date_labels = "%m/%d")+
  labs(y="Median Value")+
  nogrid

ggplot()+
  geom_line(data=daily_stats %>% filter(param=="do.mgL.mean"),aes(x=date,y=med),color=pal[1])+
  geom_ribbon(data=daily_stats %>% filter(param=="do.mgL.mean"),aes(x=date,ymin=mn,ymax=mx),fill=pal[2],alpha=0.5)+
  scale_x_date(date_breaks = "1 month",date_labels = "%m/%d")+
  theme_classic()+
  labs(x="",y="Dissolved Oxygen (mg/L)")+
  scale_y_continuous(limits=c(0,14),breaks=seq(0,14,by=2))+
  geom_hline(yintercept=c(4.3,5.0,6.0),linetype=2)+
  geom_vline(xintercept=c(as.Date("2019-02-28"),as.Date("2019-06-30"),as.Date("2019-10-31")),linetype=3,color="grey")+
  annotate("text",x=as.Date("2019-01-01"),y=4.6,label="4.3 mg/L")+
  annotate("text",x=as.Date("2019-01-01"),y=5.3,label="5.0 mg/L")+
  annotate("text",x=as.Date("2019-01-01"),y=6.3,label="6.0 mg/L")+
  ggtitle("Average Seasonal DO Concentration in Delaware River",
          subtitle="at Chester, PA, 2009-2022")
  
ggplot()+
  geom_line(data=daily_stats %>% filter(param=="posat.mean"),aes(x=date,y=med),color=pal[1])+
  geom_ribbon(data=daily_stats %>% filter(param=="posat.mean"),aes(x=date,ymin=mn,ymax=mx),fill=pal[2],alpha=0.5)+
  scale_x_date(date_breaks = "1 month",date_labels = "%m/%d")+
  theme_classic()+
  labs(x="",y="Percent Oxygen Saturation")+
  scale_y_continuous(limits=c(0,120),breaks=seq(0,120,by=20))+
  geom_hline(yintercept=c(68,77),linetype=2)+
  geom_vline(xintercept=c(as.Date("2019-02-28"),as.Date("2019-06-30"),as.Date("2019-10-31")),linetype=3,color="grey")+
  annotate("text",x=as.Date("2019-01-01"),y=80,label="77%")+
  annotate("text",x=as.Date("2019-01-01"),y=71,label="68%")+
  ggtitle("Average Seasonal Percent Oxygen Saturation in Delaware River",
          subtitle="at Chester, PA, 2009-2022")

ggplot()+
  geom_line(data=daily_stats %>% filter(param=="wt.mean"),aes(x=date,y=med),color=pal[1])+
  geom_ribbon(data=daily_stats %>% filter(param=="wt.mean"),aes(x=date,ymin=mn,ymax=mx),fill=pal[2],alpha=0.5)+
  scale_x_date(date_breaks = "1 month",date_labels = "%m/%d")+
  theme_classic()+
  labs(x="",y="Water Temperature (deg C)")+
  scale_y_continuous(limits=c(0,32),breaks=seq(0,32,by=2))+
  geom_vline(xintercept=c(as.Date("2019-02-28"),as.Date("2019-06-30"),as.Date("2019-10-31")),linetype=3,color="grey")+
  geom_hline(yintercept=28,linetype=2)+
  ggtitle("Average Seasonal Water Temperature in Delaware River",
          subtitle="at Chester, PA, 2009-2022")
  
```

