---
title: "7.02 Compare Simulated Abundance with CPUE.Rmd"
author: "J. Hagy"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(wesanderson)
library(lmodel2)
library(gt)

basic_theme <- theme_classic() +
  theme(axis.text=element_text(size=14),
        axis.title=element_text(size=14))
nogrid <- theme(panel.grid=element_blank())
largetext <- theme(axis.text=element_text(size=12),
                   axis.title = element_text(size=12),
                   strip.text=element_text(size=12))

pal <- wes_palette("Zissou1",5,type="discrete")

dataPath <- "data/"
figuresPath <- "figures/"

```

# Overview
This program loads the simulated abundance from the cohort models at Chester and Penn's Landing for 2002-2022.  It also loads the young of the year (YOY) juvenile abundance survey data obtained from the DNREC monitoring program.  The program makes a graph comparing predicted survivorship on October 31 from the model with catch per unit effort (CPUE) from the surveys.  CPUE is shown on a log scale because CPUE increased exponentially with survivorship.

I would expect that CPUE would be linearly related to % survivorship, however, the relationship for Penn's Landing was log-linear, while there was no clear relationship at all at Chester.

## Load the data sets

```{r LoadDataSets}

# Load the simulation results for Chester and Penn's Landing
load(paste0(dataPath,"5.03 Chester Cohort Simulation 2002-2022.Rdata"))
load(paste0(dataPath,"5.04 Penns Landing Cohort Simulation 2002-2022.Rdata"))

# Create variable yoy.survey with YOY survey data.
source("raw/create_yoy_survey_data.R")

```

# Review CPUE Data

```{r}

# print a table showing CPUE data so that it can be compared with original data
# for QA
gt(yoy.survey)

```

## Calculate Simulated Abundance on 10/31 and Merge with CPUE data 

```{r Calculate Simulated Abundance}

# Combine data from Chester and Penn's Landing, 
# calculate the percent survivorship on 10/31 (doy=304)
# by dividing the abundance by the initial abundance
# which was 10,000.
df <- rbind(
  sim.Chester.Zmin %>% 
  filter(doy==304) %>% # doy=304 is October 31.
  mutate(year=as.numeric(year),
         survivorship=abundance*100/10000,
         Site="Chester") %>% 
  select(Site,year,survivorship),
  sim.PennsLanding.Zmin %>% 
  filter(doy==304) %>% 
  mutate(year=as.numeric(year),
         survivorship=abundance*100/10000,
         Site="Penn's Landing") %>% 
  select(Site,year,survivorship)
)

# merge combined data with fish survey data
df <- left_join(df,yoy.survey,by="year") %>% 
  mutate(Site=factor(Site,levels=c("Penn's Landing","Chester"))) %>% 
  mutate(log.cpue=log10(yoy.cpue.1000))
```

# Fit Correlations and Model II Regressions 

```{r Fit Model II Regressions, echo=FALSE}

# Penn's Landing
df.1 <- df %>% filter(yoy.cpue.1000>0) %>%
  filter(Site=="Penn's Landing")

cor(df.1$survivorship,df.1$log.cpue,use="pairwise")^2
cor.test(df.1$survivorship,df.1$log.cpue,use="pairwise")

# Chester
df.2 <- df %>% filter(yoy.cpue.1000>0) %>%
  mutate(log.cpue=log(yoy.cpue.1000)) %>% 
  filter(Site=="Chester")

cor(df.2$survivorship,df.2$log.cpue,use="pairwise")^2
cor.test(df.2$survivorship,df.2$log.cpue,use="pairwise")

# Summary statistics for Penn's Landing
summary(select(df.1,yoy.cpue.1000,survivorship))

# Fit Model II regressions
fit <- lmodel2(log.cpue ~ survivorship,data=df.1,nperm=99,
               range.x="relative",range.y="interval")
# Get coefficients of Ranged Major Axis Regression
rma <- fit$regression.results[4,]

```

# Graph Survivorship vs. CPUE

```{r GraphSurvivorship}

dummy <- data.frame(Site=factor(c("Penn's Landing","Chester")),
                    slp=c(as.numeric(rma[3]),NA),
                    int=c(as.numeric(rma[2]),NA),
                    lab_x=c(45,NA),
                    lab_y=c(0.0017,NA),
                    lab_text=c("r2=0.56\np<0.01",NA))

plt <- df %>%
  mutate(Site=factor(Site,levels=c("Penn's Landing","Chester"))) %>% 
  filter(yoy.cpue.1000>0) %>% 
  ggplot(aes(x=survivorship,y=yoy.cpue.1000))+
  facet_wrap(~Site,ncol=1)+
  geom_point(size=5,color=pal[2])+
  geom_abline(data=dummy,aes(slope=slp,intercept=int),
                 linewidth=1,color=pal[2])+
  scale_y_log10()+
  labs(x="Simulated Cohort Survivorship (%)",y="Young-of-the-Year Catch Per Unit Effort")+
  geom_text(data=dummy,aes(x=lab_x,y=lab_y,label=lab_text))+
  nogrid
plt

# This creates Figure 4 in the TSD
pdf(file=paste0(figuresPath,"7.02 CPUE vs. Simulated Survivorship.pdf"),
    width=4, height=6, paper="letter")
  plt
dev.off()

```

# Compared expected CPUE with predicted

```{r Expected CPUE vs Predicted}

tmp <- data.frame(surv=c(1,10,30)) %>% 
  mutate(log_rma_CPUE=as.numeric(rma[2])+surv*as.numeric(surv*rma[3])) %>% 
  mutate(rma_CPUE=10^log_rma_CPUE)
tmp

```

